MG7: Configurable and scalable 16S metagenomics data analysis -- new methods optimized for massive cloud computing

### 2015-09-14 docs/paper/out/paper.tex
fix Nature News ref in intro -- paper
make minor edits in first part of intro -- paper

During the past few years, metagenomics data analysis is growing
exponentially. Some of the reasons behind this are the increasing
throughput of massively parallel sequencing technologies (with the
derived decrease in sequencing costs), and the wide impact of
metagenomics studies, especially in human health (diagnostics,
treatments, drug response or prevention). We should also mention what
could be called the microbiome explosion: all kind of microbiomes (gut,
mouth, skin, urinary tract, airway, milk, bladder) are now routinely
sequenced in different conditions of health and disease, or after
different treatments. The impact of Metagenomics is also being felt in
environmental sciences, crop sciences, the agrifood sector and
biotechnology in general. These new possibilities for exploring the
diversity of micro--organisms in the most varied environments are
opening new research areas, and drastically changing the existing ones.

@@

Genome researchers have raised the alarm over big data in the past
\href{http://www.nature.com/news/genome-researchers-raise-alarm-over-big-data-1.17912}{nature
news add ref} but even a more serious challenge might be faced with the
metagenomics boom/ upswing. If we compare metagenomics data with other
genomics data used in clinical genotyping we find a differential
feature: the key role of time. Thus, for example, in some longitudinal
studies, serial sampling of the same patient along several weeks (or
years) is being used for the follow up of some intestinal pathologies,
for studying the evolution of gut microbiome after antibiotic treatment,
or for colon cancer early detection \citep{zeller2014potential}. This
need of sampling across time adds more complexity to metagenomics data
storage and demands adapted algorithms to detect state variations across
time as well as idiosyncratic commonalities of the microbiome of each
individual \citep{franzosa2015identifying}. In addition to the
intra-individual sampling-time dependence, metagenomic clinical test
results vary depending on the specific region of extraction of the
clinical specimen. This local variability adds complexity to the
analysis since different localizations (different tissues, different
anatomical regions, healthy or tumour tissues) are required to have a
sufficiently complete landscape of the human microbiome. Moreover,
reanalysis of old samples using new tools and better reference databases
might be also demanded from time to time.

>>>>

During the past decade, metagenomics data analysis is growing
exponentially. Some of the reasons behind this are the increasing
throughput of massively parallel sequencing technologies (with the
derived decrease in sequencing costs), and the wide impact of
metagenomics studies, especially in human health (diagnostics,
treatments, drug response or prevention). We should also mention what
could be called the microbiome explosion: all kind of microbiomes (gut,
mouth, skin, urinary tract, airway, milk, bladder) are now routinely
sequenced in different conditions of health and disease, or after
different treatments. The impact of Metagenomics is also being felt in
environmental sciences, crop sciences, the agrifood sector and
biotechnology in general. These new possibilities for exploring the
diversity of micro--organisms in the most varied environments are
opening new research areas, and drastically changing the existing ones.

@@

Genome researchers have raised the alarm over big data in the past
\citep{hayden2015genome}, but even a more serious challenge might be
faced with the metagenomics boom. If we compare metagenomics data with
other genomics data used in clinical genotyping we find a differential
feature: the key role of time. Thus, for example, in some longitudinal
studies, serial sampling of the same patient along several weeks (or
years) is being used for the follow up of some intestinal pathologies,
for studying the evolution of the gut microbiome after antibiotic
treatment, or for colon cancer early detection
\citep{zeller2014potential}. This need of sampling across time adds more
complexity to metagenomics data storage and demands adapted algorithms
to detect state variations across time as well as idiosyncratic
commonalities of the microbiome of each individual
\citep{franzosa2015identifying}. In addition to the intra-individual
sampling-time dependence, metagenomic clinical test results vary
depending on the specific region of extraction of the clinical specimen.
This local variability adds complexity to the analysis since different
localizations (different tissues, different anatomical regions, healthy
or tumour tissues) are required to have a sufficiently complete
landscape of the human microbiome. Moreover, reanalysis of old samples
using new tools and better reference databases might be also demanded
from time to time.
